```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## introduction

The following exercises are modified from Chapters 3, 4, and 5 of [Geocomputation with R by Rovin Lovelace](https://r.geocompx.org/).

## prerequisites

```{r load, include=TRUE, message=FALSE, warning=FALSE}
library(sf)
library(terra)
library(spData)
library(geodata)
library(spDataLarge)
library(tidyverse)
```

```{r}
dem = rast(system.file("raster/dem.tif", package = "spDataLarge"))
landsat = rast(system.file("raster/landsat.tif", package = "spDataLarge"))
spain_dem = geodata::elevation_30s(country = "Spain", path = ".", mask = FALSE)
srtm = rast(system.file("raster/srtm.tif", package = "spDataLarge"))
```

## Exercise 1: Plot the histogram and the boxplot of the dem.tif file from the spDataLarge package

```{r dem.tif hist and boxplot}

# creating a histogram of the dem.tif file
hist(dem) # base R histogram, good for prelim data exploration

boxplot(dem) # base R boxplot

```

## Exercise 2: Manipulate rasters

reclassify the elevation in three classes: low (<300), medium and high (>500). 

```{r reclassify levels}
plot(dem)

# Inf = infinity
rcl = matrix(c(-Inf, 300, 0, # lowest value of first category, highest value of first category, category number = 0
               300, 500, 1, # lowest value of second category, highest value of second category, category number = 1
               500, Inf, 2), # lowest value of third category, highest value of third category, category number = 2 (these numbers of arbitrary, only matters for "id" vector in renaming levels)
             ncol = 3,  # 3 columns
             byrow = TRUE) # default is by column
# can't have multiple data types in one matrix i guess?

rcl

?classify
# reclassify the raster
dem_rcl <- classify(dem, rcl = rcl)

# update 0, 1, 2 to be low, medium, high
# make a dataframe where ID = 0:2
levels(dem_rcl) = data.frame(id = 0:2, cats = c("low", "medium", "high"))

# view plot
plot(dem_rcl)

```

compute the mean elevation for each altitudinal class
- zonal function

```{r zonal}
elevation_mean = zonal(dem, # first raster layer object
                       dem_rcl, # raster layer with codes representing zones, reclassified
                       fun = "mean" # function for summary stats
                       ) 

# like group by, summarize

elevation_mean

```

Calculate the Normalized Difference Water Index (NDWI; (green - nir)/(green + nir)) of a Landsat image.


```{r ndwi}
plot(landsat) # base R works well for plotting basic rasters with not too many layers

ndwi_fun = function(green, nir) {
  (green - nir) / (green + nir)
}

ndwi_rast = lapp(landsat[[c(2,  # green channel
                            4) # nir channel (layers from landsat, in documentation)
                          ]], 
                 fun = ndwi_fun) # specifying function to apply
ndwi_rast

plot(ndwi_rast)
```

calculate a correlation between NDVI and NDWI for this area

```{r}
ndvi_fun = function(nir, red){
  (nir - red) / (nir + red)
}

ndvi_rast = lapp(landsat[[c(4, 3)]], fun = ndvi_fun)
ndvi_rast
```

```{r}
combine_rast = c(ndvi_rast, ndwi_rast)

plot(combine_rast)

```


```{r}
# ?layerCor

layerCor(combine_rast, fun = cor) # using just cor (full correlation matrix between these layers)

# values indicate that they are quite similar -- almost opposite, but not quite

# correlation between the values in the matrix
```


Use `terra::distance()` to compute distances from all cells of spain to it's nearest coastline. According to the documentation, terra::distance() will calculate distance for all cells that are NA to the nearest cell that are not NA

```{r spain}
spain_dem = aggregate(spain_dem, fact = 20)
plot(spain_dem)

```

```{r spain mask plot}
# calculate distance for all cells that are NA to the nearest cell that are not NA
# calculate distance from all points to nearest coastline

# create a mask for the water

water_mask = is.na(spain_dem)

plot(water_mask) # true false mask where Spain is False, water outside is True

# anywhere that its false, update to NA for our terra::distance()
water_mask[water_mask == 0] = NA

plot(water_mask)


distance_to_coast <- terra::distance(water_mask) # default is meters

# convert to km
distance_to_coast_km <- distance_to_coast / 1000

# plot how far from the coast every point in Spain is
plot(distance_to_coast_km, main = "Distance to Coast in Spain (km)")

```


Try to modify the approach used in the above exercise by weighting the distance raster with the elevation raster; **every 100 altitudinal meters should increase the distance to the coast by 10 km.** Next, compute and visualize the difference between the raster created using the Euclidean distance (E7) and the raster weighted by elevation.

```{r}
# straight linear distance isn't super accurate (doesn't account for elevation)

distance_to_coast_km2 <- distance_to_coast_km + ((spain_dem / 100) * 10) # adds some amount of distance for every meter elevation


plot(distance_to_coast_km)
plot(distance_to_coast_km2)

# skipped Euclidean distance part

```


## Exercise 3:: Geometry Operations with Rasters

The srtm raster has a resolution of **0.00083 by 0.00083** degrees. Change its resolution to **0.01 by 0.01** degrees using all of the method available in the terra package. Visualize the results. 

```{r resample}
plot(srtm)

# extent of srtm raster, resolution of 0.01
rast_template <- rast(ext(srtm), res = 0.01)

# ?resample
srtm_resampl1 <- resample(x = srtm, 
                          y = rast_template, 
                          method = "near")

srtm_resampl2 <- resample(x = srtm, 
                          y = rast_template, 
                          method = "bilinear")
srtm_resampl3 <- resample(x = srtm, 
                          y = rast_template, 
                          method = "cubic")

srtm_resampl4 <- resample(x = srtm, 
                          y = rast_template, 
                          method = "cubicspline")
srtm_resampl5 <- resample(x = srtm, 
                          y = rast_template, 
                          method = "lanczos")
srtm_all = c(srtm_resampl1, srtm_resampl2, srtm_resampl3, srtm_resampl4, srtm_resampl5)

# viewing which method: nearest neighbor, bilinear/cubic/cubicspline interpolation, lanczos windowed sinc rsampling makes the most sense / looks the best
plot(srtm_all)
```

```{r}
srtm_resampl_all = c(srtm_resampl1, srtm_resampl2, srtm_resampl3,
                     srtm_resampl4, srtm_resampl5)
plot(srtm_resampl_all)
```

Please take this [survey](https://forms.gle/P6EneoEW7GY56LP16) to give me feedback on last week and this week!


```{r}
layerCor(srtm_all, fun = cor)
```

