---
title: 'EDS 223: week 4 lab'
author: "Ruth Oliver"
date: "2022-10-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## introduction
In this lab, we'll explore the basics of working with raster data, including attribute, spatial, and geometry operations. This lab follows chapters [3](https://geocompr.robinlovelace.net/attr.html), [4](https://geocompr.robinlovelace.net/spatial-operations.html), and [5](https://geocompr.robinlovelace.net/geometry-operations.html) of Geocomputation with R by Robin Lovelace.

## prerequisites

```{r install, include=TRUE}
library(terra)
library(dplyr)
library(spData)
library(spDataLarge)
library(tmap)
library(geodata)
```

- in sf, we called things sf objects
- in terra, we'll call them spat rasters
## manipulating raster objects
Raster data represents continuous surfaces, as opposed to the discrete features represented in the vector data model. Here we'll learn how to create raster data objects from scratch and how to do basic data manipulations.

Let's create a **SpatRaster** object using a digitial elevation model for Zion National Park.

```{r include=TRUE}
raster_filepath <- system.file("raster/srtm.tif", # zion raster specifically
                               package = "spDataLarge") # getting this into our environment

my_rast <- rast(raster_filepath)

# class(my_rast)

plot(my_rast)
```

We can also create rasters from scratch using the **rast()** function. Here we create 36 cells centerd around (0,0). By default the CRS is set to WGS84, but we could change this with the "crs" argument.
Because we are working in WGS84, the resolution is in units of degrees. **rast()** fills the values of the cells row-wise starting in the upper left corner.


```{r include=TRUE}
# crs by default is wgs 84
new_raster <- rast(nrow = 6, ncol = 6, resolution = 0.5,
                   xmin = -1.5, xmax = 1.5,
                   ymin = -1.5, ymax = 1.5,
                   vals = 1:36,
                   crs = "EPSG:5070")
new_raster

plot(new_raster)
```

The SpatRaster class can also handle multiple layers.

```{r include=TRUE}
# load in raster w multiple layers
multi_raster_file <- system.file("raster/landsat.tif", package = "spDataLarge")

multi_rast <- rast(multi_raster_file)

multi_rast

plot(multi_rast)
```

We can subset layers using either the layer number or name
```{r include=TRUE}
multi_rast_3 <- subset(multi_rast, 3)

multi_rast_3

multi_rast_4 <- subset(multi_rast, "landsat_4")
multi_rast_4


```

We can combine SpatRaster objects into one, using the c function
```{r include=TRUE}
multi_rast_3 <- subset(multi_rast, 3) # can select multiple by doing c(3, 4)
multi_rast_4 <- subset(multi_rast, "landsat_4")

multi_rast_34 <-  c(multi_rast_3, multi_rast_4)
multi_rast_34
```

Let's create an example raster for elevation

```{r include=TRUE}
elev <- rast(nrows = 6, ncols = 6, resolution = 0.5,
             xmin = -1.5, xmax = 1.5,
             ymin = -1.5, ymax = 1.5,
             vals = 1:36)

grain_order <- c("clay", "silt", "sand")

# randomly create a vector of 36 of these attributes

grain_character <- sample(grain_order, 36, replace = TRUE)
# grain_character

# taking character string and turning it into factors
grain_fact <- factor(grain_character, levels = grain_order)
# grain_fact

# let's put this into a raster!

grain <- rast(nrows = 6, ncols = 6, resolution = 0.5,
             xmin = -1.5, xmax = 1.5,
             ymin = -1.5, ymax = 1.5,
             vals = grain_fact # vector of factors of soil types
             )

# grain

plot(grain)

plot(elev)

```

Rasters can also hold categorical data. Let's create an example raster for soil types.

```{r include=TRUE}

# did above

```



## raster subsetting
We can index rasters using, row-column indexing, cell IDs, coordinates, other spatial objects.

```{r include=TRUE}
# row 1, column 1

elev[1, 2]

elev[1] # gives you the value in the first cell

elev[36] # 36th cell




```

If we had a two layered raster, subsetting would return the values in both layers.

```{r include=TRUE}

two_layers <- c(grain, elev)

two_layers[1] # by default, the layer names are lyr.1

```

We can also modify/overwrite cell values.
```{r include=TRUE}

elev[1, 1] = 0

elev[1,1]

elev[] # takes it from a matrix to a vector (unwraps it?)

```

Replacing values in multi-layer rasters requires a matrix with as many columns as layers and rows as replaceable cells.
```{r include=TRUE}
# 1 row by 2 columns
cbind(c(1), c(4))

two_layers[1] <- cbind(c(1), c(4))

two_layers[]
```

## summarizing raster objects
We can get info on raster values just by typing the name or using the summary function.
```{r include=TRUE}

summary(elev)

```

We can get global summaries, such as standard deviation.
```{r include=TRUE}
global(elev, sd) # global used w spatRaster

```

Or we can use **freq()** to get the counts with categories.
```{r include=TRUE}
freq(grain)
hist(grain)
```

## spatial subsetting
We can move from subsetting based on specific cell IDs to extract info based on spatial objects.

To use coordinates for subsetting, we can "translate" coordinates into a cell ID with the **terra** function **cellFromXY()** or **terra::extract()**.

```{r include=TRUE}
# if we want to know the value at a certain point, make coords, translate to cell ID, extract info from those cells

id <- cellFromXY(elev, xy = matrix(c(0.1, 0.1), ncol = 2))
# gives you the ID of the cell that it at that point in spce
elev[id]

# extract is doing what the previous 2 steps did in one
terra::extract(elev, matrix(c(0.1, 0.1), ncol = 2))

# gets ID, pulls out the value

```


Raster objects can also subset with another raster object. Here we extract the values of our elevation raster that fall within the extent of a masking raster.

```{r include=TRUE}
# clipping
# define a raster to clip to
clip <- rast(xmin = 0.9, xmax = 1.8,
             ymin = -0.45, ymax = 0.45,
             resolution = 0.3, # we don't have to specify the rows and columns bc resolution automatically does area divided by # of boxes -- 9 boxes, repeat 1 nine times
             vals = rep(1,9))

test <- elev[clip]
# clipped to the area, pulled out the cells within it
# give me the cells that overlap w the clip raster
# we lost the geometry, just have the values, don't have it as a raster anymore
test

test2 <- terra::extract(elev, ext(clip)) # subset it to the extent of the clip raster
test2
# not sure why this just returns NULL, come back after Ruth tells us later :)
```

In the previous example, we just got the values back. In some cases, we might want the output to be the raster cells themselves. We can do this use the "[" operator and setting "drop = FALSE"
- dropped the geometries -- what if we want them!?

This example returns the first 2 cells of the first row of the "elev" raster.

```{r include=TRUE}
elev[1:2, drop = FALSE]

```

Another common use of spatial subsetting is when we use one raster with the same extent and resolution to mask the another. In this case, the masking raster needs to be composed of logicals or NAs.

```{r mask, include=TRUE}
# resolution, extent, and position have to match
rmask <- elev
values(rmask) <- sample(c(NA, TRUE), 36, replace = TRUE)
rmask
plot(rmask)


# subset elevation raster to this mask

test3 <- elev[rmask]
#test3

# plot(rmask)
test3
# plot(test3)


test4 <- elev[rmask, drop = FALSE]
#test4
plot(test4) # didn't work...

test5 <- mask(elev, rmask) # the terra version works!
test5

plot(test5)

summary(test5)

```

We can also use a similar approach to replace values that we suspect are incorrect.

```{r include=TRUE}

elev[elev < 20] = NA

plot(elev)



```

## map algebra
Here we define map algebra as the set of operations that modify or summarize raster cell values with reference to surrounding cells, zones, or statistical functions that apply to every cell.

## local operations
Local operations are computed on each cell individually. We can use oridinary arithemetic or logical statements.

```{r include=TRUE}
elev + elev
# created a new raster that is just doubling the values in every cell

elev^2
# raised them to the power of 2

# log :)
log(elev)

elev > 5 # filter, returns logical 
```

We can also classify intervals of values into groups. For example, we could classify a DEM into low, middle, and high elevation cells

-   first we need to construct a reclassification matrix
-   the first column corresponds to the lower end of the class
-   the second column corresponds to the upper end of the class
-   the third column corresponds to the new value for the specified ranges in columns 1 and 2

```{r include=TRUE}


```

For more efficient processing, we can use a set of map algebra functions.
-   **app()** applies a function to each cell of a raster to summarize the values of multiple layers into one layer
-   **tapp()** is an extension of "app()" that allows us to apply on operation on a subset of layers
-   **lapp()** allows us to apply a function to each cell using layers as arguments

We can use the **lapp()** function to compute the Normalized Difference Vegetation Index (NDVI).
Let's calculate NDVI for Zion National Park using multispectral satellite data.

```{r include=TRUE}
# creating our own NDVI layer
# can look at reflection of _infrared__ to figure out what material is being sensed
multi_rast




```

We need to define a function to calculate NDVI.
```{r include=TRUE}
ndvi_fun <- function(nir, red){
  (nir - red) / (nir + red)
}

```

So now we can use **lapp()** to calculate NDVI in each raster cell. To do so, we just need the NIR and red bands.

```{r include=TRUE}
# apply the function
ndvi_rast <- lapp(multi_rast[[c(3, 4)]], fun = ndvi_fun)
# for every value, we subtracted the value from cell 4 from cell 3, then divided by their sum


plot(ndvi_rast)

```


## focal operations
Local operations operate on one cell, though from multiple layers. Focal operations take into account a central (focal) cell and its neighbors. The neighborhood (or kernel, moving window, filter) can take any size or shape. A focal operation applies an aggregation function to all cells in the neighborhood and updates the value of the central cell before moving on to the next central cell

We can use the **focal()** function to perform spatial filtering. We define the size, shape, and weights of the moving window using a matrix. Here we find the minimum.

```{r include=TRUE}
# define the size, shape, and weights of the moving window -- what the neighborhood we want to use is, what op. we want 
r_focal <- focal(elev, w = matrix(1, nrow = 3, ncol = 3), fun = min)
r_focal

plot(elev)
plot(r_focal)
```

## zonal operations
Similar to focal operations, zonal operations apply an aggregation function to multiple cells. However, instead of applying operations to neighbors, zonal operations aggregate based on "zones". Zones can are defined using a categorical raster and do not necessarily have to be neighbors

For example, we could find the average elevation for different soil grain sizes.
```{r include=TRUE}
plot(grain)
# kind of look like zones
# not spatially continuous -- dist. isn't cont.

test6 <- zonal(elev, grain, fun = mean, na.rm = TRUE)
zonal(elev, grain, fun = mean, na.rm = TRUE)

# it's not a raster anymore
class(test6)

```

## merging rasters
In some cases, data for a region will be stored in multiple, contiguous files. To use them as a single raster, we need to merge them.\

In this example, we download elevation data for Austria and Switzerland and merge the two rasters into one.

```{r include=TRUE}


```



## geometric operations
When merging or performing map algebra, rasters need to match in their resolution, projection, origin, and/or extent

In the simplest case, two images differ only in their extent. Let's start by increasing the extent of a elevation raster.

```{r include=TRUE}


```

Performing algebraic operations on objects with different extents doesn't work.
```{r include=TRUE, eval=FALSE}

```

We can align the extent of the 2 rasters using the extend() function. Here we extend the elev object to the extent of elev_2 by adding NAs.

```{r include=TRUE}

```

the origin function returns the coordinates of the cell corner closes to the coordinates (0,0). We can also manually change the origin.

```{r include=TRUE}

```

## aggregation and disaggregation
Faster datasets can also differ in their resolution to match resolutions we can decrease the resolution by aggregating or increase the resolution by disaggregating.

Let's start by changing the resolution of a DEM by a factor of 5, by taking the mean.

```{r include=TRUE}

```

We have some choices when increasing the resolution. Here we try the bilinear method.
```{r include=TRUE}

```

## resampling

Aggregation/disaggregation work when both rasters have the same origins. What do we do in the case where we have two or more rasters with different origins and resolutions? Resampling computes values for new pixel locations based on custom resolutions and origins.

In most cases, the target raster would be an object you are already working with, but here we define a target raster.

```{r include=TRUE}

```




