```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## introduction

The following exercises are modified from Chapters 3, 4, and 5 of [Geocomputation with R by Rovin Lovelace](https://r.geocompx.org/).

## prerequisites

```{r load, include=TRUE, message=FALSE, warning=FALSE}
library(sf)
library(terra)
library(spData)
library(geodata)
library(spDataLarge)
library(tidyverse)
```

```{r}
dem = rast(system.file("raster/dem.tif", package = "spDataLarge"))
landsat = rast(system.file("raster/landsat.tif", package = "spDataLarge"))
spain_dem = geodata::elevation_30s(country = "Spain", path = ".", mask = FALSE)
srtm = rast(system.file("raster/srtm.tif", package = "spDataLarge"))
```

## Exercise 1: Plot the histogram and the boxplot of the dem.tif file from the spDataLarge package

```{r dem.tif hist and boxplot}

# creating a histogram of the dem.tif file
hist(dem) # base R histogram, good for prelim data exploration

boxplot(dem) # base R boxplot

```

## Exercise 2: Manipulate rasters

reclassify the elevation in three classes: low (<300), medium and high (>500). 

```{r reclassify levels}
plot(dem)

# Inf = infinity
rcl = matrix(c(-Inf, 300, 0, # lowest value of first category, highest value of first category, category number = 0
               300, 500, 1, # lowest value of second category, highest value of second category, category number = 1
               500, Inf, 2), # lowest value of third category, highest value of third category, category number = 2 (these numbers of arbitrary, only matters for "id" vector in renaming levels)
             ncol = 3,  # 3 columns
             byrow = TRUE) # default is by column
# can't have multiple data types in one matrix i guess?

rcl

?classify
# reclassify the raster
dem_rcl <- classify(dem, rcl = rcl)

# update 0, 1, 2 to be low, medium, high
# make a dataframe where ID = 0:2
levels(dem_rcl) = data.frame(id = 0:2, cats = c("low", "medium", "high"))

# view plot
plot(dem_rcl)

```

compute the mean elevation for each altitudinal class
- zonal function

```{r zonal}
elevation_mean = zonal(dem, # first raster layer object
                       dem_rcl, # raster layer with codes representing zones, reclassified
                       fun = "mean" # function for summary stats
                       ) 

# like group by, summarize

elevation_mean

```

Calculate the Normalized Difference Water Index (NDWI; (green - nir)/(green + nir)) of a Landsat image.


```{r ndwi}
plot(landsat) # base R works well for plotting basic rasters with not too many layers

ndwi_fun = function(green, nir) {
  (green - nir) / (green + nir)
}

ndwi_rast = lapp(landsat[[c(2,  # green channel
                            4) # nir channel (layers from landsat, in documentation)
                          ]], 
                 fun = ndwi_fun) # specifying function to apply
ndwi_rast

plot(ndwi_rast)
```

calculate a correlation between NDVI and NDWI for this area

```{r}
ndvi_fun = function(nir, red){
  (nir - red) / (nir + red)
}

ndvi_rast = lapp(landsat[[c(4, 3)]], fun = ndvi_fun)
ndvi_rast
```

```{r}

```

Use `terra::distance()` to compute distances from all cells of spain to it's nearest coastline. According to the documentation, terra::distance() will calculate distance for all cells that are NA to the nearest cell that are not NA

```{r}
spain_dem = aggregate(spain_dem, fact = 20)
plot(spain_dem)

```

Try to modify the approach used in the above exercise by weighting the distance raster with the elevation raster; every 100 altitudinal meters should increase the distance to the coast by 10 km. Next, compute and visualize the difference between the raster created using the Euclidean distance (E7) and the raster weighted by elevation.

```{r}

```


## Exercise 3:: Geometry Operations with Rasters

The srtm raster has a resolution of 0.00083 by 0.00083 degrees. Change its resolution to 0.01 by 0.01 degrees using all of the method available in the terra package. Visualize the results. 

```{r}
plot(srtm)

```

```{r}
srtm_resampl_all = c(srtm_resampl1, srtm_resampl2, srtm_resampl3,
                     srtm_resampl4, srtm_resampl5)
plot(srtm_resampl_all)
```

Please take this [survey](https://forms.gle/P6EneoEW7GY56LP16) to give me feedback on last week and this week!
